---
title: "ST492 Final Project"
author: "Scott McGrouther"
date: "2024-03-03"
output: html_document
---

# 0) Prerequisites

# Imports

```{r}

library("readxl")
library("tidyverse")
library("TSA")

install.packages("forecast")
library("forecast")

install.packages("tseries")
library(tseries)

search()

```

# Importing the Data Set

```{r}

data <- read_excel("C:/Users/scott/OneDrive/Desktop/SEMESTER 2/ST 492/Project/Project Data - ST 492!.xlsx")

data

```

# Variables and TS Creation

```{r}

# Get Variables
time_all <- data$Date
prices_all <- data$Average

length(time_all)
#time_all

print("-------------------------------------------------------")

#prices_all

time <- time_all[1:105]
prices <- prices_all[1:105]

print("-------------------------------------------------------")

#time

print("-------------------------------------------------------")

#prices

print("-------------------------------------------------------")

# Make a Time Series Variable

prices_ts <- ts(prices, frequency = 52, start = c(2018, 1))

prices_ts

print("-------------------------------------------------------")

prices_ts_all <- ts(prices_all, frequency = 52, start = c(2018, 1))

prices_ts_all

print("-------------------------------------------------------")

length(prices_all) 
length(prices)

```

# Validation

```{r}

# Time Series ACF's Trimmed Data

acf(prices_ts)

acf(diff(prices_ts))

acf(diff(prices_ts, lag = 52))

# Time Series ACF's Entire Data

acf(prices_ts_all)

acf(diff(prices_ts_all))

acf(diff(prices_ts_all, lag = 52))

```

# Understanding the Data

```{r}

# Understaning the COVID-19 Spike

max(prices_ts_all)
min(prices_ts_all)

# Plotting the Data

plot(prices_ts, type = "o", xlab = "Year", ylab = "Price", main = "Air Canada Prices Over Time Before COVID-19 Spike")

points(y=prices_ts, x=time(prices_ts), pch = as.vector(season(prices_ts)), col=4, cex = 0.8)

plot(prices_ts_all, type = "o", xlab = "Year", ylab = "Price", main = "Air Canada Prices Over Time with COVID-19 Spike")

points(y=prices_ts_all, x=time(prices_ts_all), pch = as.vector(season(prices_ts_all)), col=4, cex = 0.8)

```






# 1) Model Identification

# Regression Approach

# Decomposition

```{r}

# Modeling Trend 

decomposition <- decompose(prices_ts)

plot(decomposition)

total <- diff(range(prices_ts))

trend <- diff(range(decomposition$trend, na.rm=T))
seasonal <- diff(range(decomposition$seasonal, na.rm=T))
random <- diff(range(decomposition$random, na.rm=T))

print("Total")
print(total)
print(paste(total/total * 100,"%"))
print("--------------------------------------")
print("Trend")
print(trend)
print(paste(trend/total * 100,"%"))
print("--------------------------------------")
print("Seasonal")
print(seasonal)
print(paste(seasonal/total * 100,"%"))
print("--------------------------------------")
print("Random")
print(random)
print(paste(random/total * 100,"%"))

```

# Trend Models

```{r}

# We Will Fit the First Five Polynomial Regression Models and See if There is Significant Differences Between Them Before Deciding


ti <- as.vector(time(prices_ts))

# Linear Model - RED

reg1 <- lm(prices_ts ~ ti)

summary(reg1)
plot(prices_ts, main = "Air Canada Stock Prices Over Time", type = "o")
lines(x=ti, y=predict(reg1), lty=2, col=2)

print("----------------------------------------------------------------------")


# Quadratic Model

reg2 <- lm(prices_ts ~ ti + I(ti^2))

plot(prices_ts, main = "Air Canada Stock Prices Over Time", type = "o")
lines(x=ti, y=predict(reg2), lty=2, col="Green")
summary(reg2)

print("----------------------------------------------------------------------")


# Cubic Model

reg3 <- lm(prices_ts ~ ti + I(ti^2) + I(ti^3))

plot(prices_ts, main = "Air Canada Stock Prices Over Time", type = "o")
lines(x=ti, y=predict(reg3), lty=2, col="Blue")
summary(reg3)

print("----------------------------------------------------------------------")


# Quartic Model

reg4 <- lm(prices_ts ~ ti + I(ti^2) + I(ti^3) + I(ti^4))

plot(prices_ts, main = "Air Canada Stock Prices Over Time", type = "o")
lines(x=ti, y=predict(reg4), lty=2, col="Orange")
summary(reg4)

print("----------------------------------------------------------------------")


# Quintic Model

reg5 <- lm(prices_ts ~ ti + I(ti^2) + I(ti^3) + I(ti^4) + I(ti^5))

plot(prices_ts, main = "Air Canada Stock Prices Over Time", type = "o")
lines(x=ti, y=predict(reg5), lty=2, col="Orange")
summary(reg5)


```


# Seasonality Models

```{r}

# Weekly

weeks <- season(prices_ts)

model_season <- lm(prices_ts ~ weeks)

summary(model_season)

plot(prices_ts, xlab = "Week", ylab = "Air Canada Stock Prices Over Time", type = "p")

lines(ti, predict(model_season), lty = 2, col = 2)


```

# Combining Both

```{r}

model_regression <- lm(prices_ts ~ weeks + ti + I(ti^3))

summary(model_regression)

plot(prices_ts, xlab = "Week", ylab = "Air Canada Stock Prices Over Time", type = "p")

lines(ti, predict(model_regression), lty = 2, col = 2)

# Not worth all of the variables

```





# ARIMA Approach

# Normal TS

```{r}

# Checking Stationary

# Recall

plot(prices_ts, type = "o", xlab = "Year", ylab = "Price", main = "Air Canada Prices Over Time Before COVID-19 Spike")

acf(prices_ts)

pacf(prices_ts)

eacf(prices_ts)

plot(armasubsets(prices_ts, nar=12, nma=12))

adf_result <- adf.test(prices_ts, alternative = "stationary", k = 1)

print(adf_result)

# Obviously NOT Stationary - First We Attempt the Box Cox

BC <- BoxCox.ar(prices_ts, lambda=seq(-2,2,0.1))

print(BC$mle)

print(BC$ci)

```

# Differenced TS

```{r}

# Difference

diff_series <- diff(prices_ts)

plot(diff_series, type = "o", xlab = "Year", ylab = "Price", main = "Air Canada Prices Over Time Before COVID-19 Spike")

acf(diff_series)

pacf(diff_series)

eacf(diff_series)

plot(armasubsets(diff_series, nar=12, nma=12))

adf_result2 <- adf.test(diff_series, alternative = "stationary", k = 1)

print(adf_result2)

```


# Log Return Series

```{r}

# Additional Test - Based on FM Theory

log_return <- diff(log(prices_ts))

plot(log_return, type = "o", xlab = "Year", ylab = "Price", main = "Air Canada Prices Over Time Before COVID-19 Spike")

acf(log_return)

pacf(log_return)

eacf(log_return)

plot(armasubsets(log_return, nar=12, nma=12))

adf_result5 <- adf.test(log_return, alternative = "stationary", k = 1)

print(adf_result5)


```








# 2) Model Fitting

# Regression Approach

# FINAL REGRESSION MODEL

```{r}

# Thus, We Obtain:

REG <- lm(prices_ts ~ ti + I(ti^3))

summary(REG)

plot(prices_ts, xlab = "Week", ylab = "Air Canada Stock Prices Over Time", type = "p")

lines(ti, predict(REG), lty = 2, col = 2)

```







# ARIMA Approach

# FINAL ARIMA MODELS

```{r}

# Thus, We Obtain:

IMA <- arima(x = prices_ts, order = c(0,1,1))

ARI <- arima(x = prices_ts, order = c(1,1,0))

IMA_2 <- arima(x = prices_ts, order = c(0,1,2))

ARI_2 <- arima(x = prices_ts, order = c(2,1,0))

ARIMA <- arima(x = prices_ts, order = c(1,1,1))

SIMA <- arima(x = prices_ts, order = c(0,1,0), seasonal=list(order=c(0,0,1), period=12))

SIMA_2 <- arima(x = prices_ts, order = c(0,1,0), seasonal=list(order=c(0,0,2), period=12))

SIMA_3 <- arima(x = prices_ts, order = c(0,1,0), seasonal=list(order=c(0,0,3), period=12))

LOG_IMA <- arima(x = log(prices_ts), order = c(0,1,1))

LOG_IMA_2 <- arima(x = log(prices_ts), order = c(0,1,2))

LOG_ARI <- arima(x = log(prices_ts), order = c(1,1,0))

LOG_ARI_2 <- arima(x = log(prices_ts), order = c(2,1,0))

LOG_ARIMA <- arima(x = log(prices_ts), order = c(1,1,1))

LOG_SIMA <- arima(x = log(prices_ts), order = c(0,1,0), seasonal=list(order=c(0,0,1), period=12))

LOG_SIMA_2 <- arima(x = log(prices_ts), order = c(0,1,0), seasonal=list(order=c(0,0,2), period=12))

LOG_SIMA_3 <- arima(x = log(prices_ts), order = c(0,1,0), seasonal=list(order=c(0,0,3), period=12))

IMA

print("---------------------------------------------------------------------------------------------------")

IMA_2

print("---------------------------------------------------------------------------------------------------")

ARI

print("---------------------------------------------------------------------------------------------------")

ARI_2

print("---------------------------------------------------------------------------------------------------")

ARIMA

print("---------------------------------------------------------------------------------------------------")

SIMA

print("---------------------------------------------------------------------------------------------------")

SIMA_2

print("---------------------------------------------------------------------------------------------------")

SIMA_3

print("---------------------------------------------------------------------------------------------------")

LOG_IMA

print("---------------------------------------------------------------------------------------------------")

LOG_IMA_2

print("---------------------------------------------------------------------------------------------------")

LOG_ARI

print("---------------------------------------------------------------------------------------------------")

LOG_ARI_2

print("---------------------------------------------------------------------------------------------------")

LOG_ARIMA

print("---------------------------------------------------------------------------------------------------")

LOG_SIMA

print("---------------------------------------------------------------------------------------------------")

LOG_SIMA_2

print("---------------------------------------------------------------------------------------------------")

LOG_SIMA_3

```





# 3) Model Checking

# Regression Approach

# CHECKING THE MODEL - REG

```{r}

# Recall:

summary(REG)

plot(prices_ts, xlab = "Week", ylab = "Air Canada Stock Prices Over Time", type = "p")

lines(ti, predict(REG), lty = 4, col = 2)

# Needed

residual_REG <- residuals(REG)
predicted_REG <- predict(REG)


# Normality

dev.new(width=8, height = 4)
hist(residual_REG, main = "Air Canada Stock: Residuals", xlab = "Residuals")
qqnorm(residual_REG)
qqline(residual_REG)
shapiro.test(residual_REG)

# Constant Variance

plot(predicted_REG, residual_REG, main = "Fitted Versus Residuals" , xlab = "Fitted", ylab = "Residuals")

# White Noise

plot(residual_REG, ylab = "Residuals", main = "Residuals" )
acf(residual_REG, main = "ACF of Residuals")

# Runs Test of Independence

print("----------------------------------------------------------------------")

runs(residual_REG)

```


# ARIMA Approach

# Function for a General Model 

```{r}

residual_analysis <- function(model, prices_ts, ti){
  
  # Needed
  
  residual <- rstandard(model) # Residual
  rez <- residuals(model)
  
  fitted_model <- prices_ts - rez

  # Plotting
  
  plot(prices_ts, xlab = "Week", ylab = "Air Canada Stock Prices Over Time", type = "p")
  lines(ti, fitted_model, lty = 4, col = 2)
  
  # Normality
  
  hist(residual, main = "Air Canada Stock: Residuals", xlab = "Residuals")
  qqnorm(residual)
  qqline(residual)
  print(shapiro.test(residual))
  
  # Constant Variance
  
  print(as.vector(fitted_model))
  print(length(as.vector(fitted_model)))
  print(length(as.vector(residual)))
  
  plot(as.vector(fitted_model), as.vector(residual), main = "Fitted Versus Residuals" , xlab = "Fitted", ylab = "Residuals")
  
  # White Noise
  
  plot(residual, ylab = "Residuals", main = "Residuals" )
  acf(residual, main = "ACF of Residuals")
  
  
  # Ljung-Box Test
  
  
  LBpvals <- rep(NA, 15)
  
  for(i in 5:15){
    LBpvals[i] <- LB.test(model, lag = i)$p.value
    
  }
  
  plot(LBpvals, ylim = c(0,1), main = "P-Values from LB Test", ylab = "P-Values", xlab = "Lag")
  abline(h = 0.05, lty = 2, col = "red")
  
  
  # Runs Test of Independence
  
  print(runs(residual))
  
  # At Once - HARD TO SEE
  
  tsdiag(model)
  
  
}

```

# IMA(0,1,1) - IMA

```{r}

residual_analysis(IMA, prices_ts, ti)

```
# ARI(1,1,0) - ARI

```{r}

ARI
residual_analysis(ARI, prices_ts, ti)

```

# SARIMA(0,1,1)_12 - SIMA

```{r}

SIMA
residual_analysis(SIMA, prices_ts, ti)

```

# Log Return Model - IMA

```{r}

LOG_IMA
residual_analysis(LOG_IMA, prices_ts, ti)

```

# Log Return Model - ARI

```{r}

LOG_ARI
residual_analysis(LOG_ARI, prices_ts, ti)

```

# Log SARIMA(0,1,1)_12 - LOG_SIMA

```{r}

LOG_SIMA
residual_analysis(LOG_SIMA, prices_ts, ti)

```



# 4) Predictions

# Chosen Models

# - REGRESSION

```{r}

REG

print("--------------------------------------------------------------")

summary(REG)

# REGRESSION FUNCTION 

REGRESSION <- function(REG, new_data){
  
  coefficients <- coef(REG)
  
  Y <- coefficients[1] + coefficients[2] * new_data + coefficients[3]*new_data^3
  
  return(Y)
  
} 

# Creating a Zoom in Vector

prices_ts_visual <- ts(prices_all[100:157], frequency = 52, start = c(2019, 48))

ti_all <- as.vector(time(prices_ts_all))
ti_visual <- as.vector(time(prices_ts_visual))


# Seeing the Forecast On Training Data

plot(prices_ts, type = "l", ylim = c(10,100),
     xlab = "Time", ylab = "Air Canada Stock Prices",
     main = "Regression Model over Training Data Time Period")

lines(ti, REGRESSION(REG, ti), lty = 4, col = 2)
legend("topleft", legend = c("Regression"),
       col = c("red"), lty = c(4))
# Normal Data

plot(prices_ts_all, type = "l", ylim = c(10,100),
     xlab = "Time", ylab = "Air Canada Stock Prices",
     main = "Regression Model over Entire Data Time Period")

lines(ti_all, REGRESSION(REG, ti_all), lty = 4, col = 2)
legend("topleft", legend = c("Regression"),
       col = c("red"), lty = c(4))

# Zoomed

plot(prices_ts_visual, type = "l", ylim = c(10,100),
     xlab = "Time", ylab = "Air Canada Stock Prices",
     main = "Regression Model over Zoomed Data Time Period")

lines(ti_visual, REGRESSION(REG, ti_visual), lty = 4, col = 2)
legend("topleft", legend = c("Regression"),
       col = c("red"), lty = c(4))


```

# - IMA

```{r}

IMA

# Formula

prices_ts[105] + 0.3242*rstandard(IMA)[105]

# Getting Values

IMA_predictions <- predict(IMA, n.ahead = 52, se.fit = TRUE)

predicted_values <- IMA_predictions$pred
standard_errors <- IMA_predictions$se

upper_bound <- predicted_values + 1.96 * standard_errors
lower_bound <- predicted_values - 1.96 * standard_errors

#predicted_values
#upper_bound
#lower_bound

# Seeing the Forecast On Training Data

plot(prices_ts, type = "l", ylim = c(10,85),
     xlab = "Time", ylab = "Air Canada Stock Prices",
     main = "Next 52 Predictions with 95% Prediction Intervals")

lines(predicted_values, col = "blue", lty = 1)
lines(upper_bound, col = "red", lty = 2)
lines(lower_bound, col = "red", lty = 2)

legend("topright", legend = c("Predicted", "95% Prediction Interval"),
       col = c("blue", "red"), lty = c(1, 2))



# Normal Data

plot(prices_ts_all, type = "l", ylim = c(10,85),
     xlab = "Time", ylab = "Air Canada Stock Prices",
     main = "Next 52 Predictions with 95% Prediction Intervals, On Training Dataset")

lines(predicted_values, col = "blue", lty = 1)
lines(upper_bound, col = "red", lty = 2)
lines(lower_bound, col = "red", lty = 2)

legend("topright", legend = c("Predicted", "95% Prediction Interval, On Entire Dataset"),
       col = c("blue", "red"), lty = c(1, 2))


# Zoomed

plot(prices_ts_visual, type = "l", ylim = c(10,85),
     xlab = "Time", ylab = "Air Canada Stock Prices",
     main = "Next 52 Predictions with 95% Prediction Intervals, On Zoomed Dataset")

lines(predicted_values, col = "blue", lty = 1)
lines(upper_bound, col = "red", lty = 2)
lines(lower_bound, col = "red", lty = 2)

legend("topright", legend = c("Predicted", "95% Prediction Interval"),
       col = c("blue", "red"), lty = c(1, 2))


```

# - ARI

```{r}

ARI

# Getting Values

ARI_predictions <- predict(ARI, n.ahead = 52, se.fit = TRUE)

predicted_values <- ARI_predictions$pred
standard_errors <- ARI_predictions$se

upper_bound <- predicted_values + 1.96 * standard_errors
lower_bound <- predicted_values - 1.96 * standard_errors

#predicted_values
#upper_bound
#lower_bound

# Seeing the Forecast On Training Data

plot(prices_ts, type = "l", ylim = c(10,85),
     xlab = "Time", ylab = "Air Canada Stock Prices",
     main = "Next 52 Predictions with 95% Prediction Intervals, On Training Dataset")

lines(predicted_values, col = "blue", lty = 1)
lines(upper_bound, col = "red", lty = 2)
lines(lower_bound, col = "red", lty = 2)

legend("topright", legend = c("Predicted", "95% Prediction Interval"),
       col = c("blue", "red"), lty = c(1, 2))



# Normal Data

plot(prices_ts_all, type = "l", ylim = c(10,85),
     xlab = "Time", ylab = "Air Canada Stock Prices",
     main = "Next 52 Predictions with 95% Prediction Intervals, On Entire Dataset")

lines(predicted_values, col = "blue", lty = 1)
lines(upper_bound, col = "red", lty = 2)
lines(lower_bound, col = "red", lty = 2)

legend("topright", legend = c("Predicted", "95% Prediction Interval"),
       col = c("blue", "red"), lty = c(1, 2))


# Zoomed

plot(prices_ts_visual, type = "l", ylim = c(10,85),
     xlab = "Time", ylab = "Air Canada Stock Prices",
     main = "Next 52 Predictions with 95% Prediction Intervals, On Zoomed Dataset")

lines(predicted_values, col = "blue", lty = 1)
lines(upper_bound, col = "red", lty = 2)
lines(lower_bound, col = "red", lty = 2)

legend("topright", legend = c("Predicted", "95% Prediction Interval"),
       col = c("blue", "red"), lty = c(1, 2))


```